---
title: "tidyr"
author: "Helen Fredricks"
date: "October 23, 2018"
output: html_document
---

TIDYR cleaning datasets - the goal is long data, not wide data... 

```{r}
## read in gapminder data in wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')
```

gather / spread

tidyr::gather()

Gather columns into key-value pairs.

Description

Gather takes multiple columns and collapses into key-value pairs, duplicating all other columns as needed. You use gather() when you notice that you have columns that are not variables.

Usage

gather(data, key = "key", value = "value", ..., na.rm = FALSE,
  convert = FALSE, factor_key = FALSE)


```{r}
#gap_long the new dataset

gap_long <- gap_wide %>% 
  gather(key = obstype_year,
         value = obs_values,
         -continent, -country)


# don't need 
gap_long <- gap_wide %>% 
  gather(obstype_year,
         obs_values,
         -continent, -country)
```

seperate

```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year,
         value = obs_values,
         -continent, -country) %>% 
  separate(col = obstype_year,
           into = c("obs_type", "year"),
           sep = "_",
           convert = TRUE)
#convert will format numbers if they are numbers, 

# check the structure of the dataframe
str(gap_long)
```

plot long data

```{r}
canada_df <- gap_long %>% 
  filter(country == "Canada",
         obs_type == "lifeExp")

ggplot(data = canada_df, aes(x = year, y = obs_values)) +
  geom_line()


```

collect the americas data

```{r}
life_df <- gap_long %>% 
  filter(obs_type == "lifeExp",
         continent == "Americas")

ggplot(data = life_df, aes(x = year, y = obs_values, colour = country)) +
       geom_line()
```

7.5 Exercise

    Using gap_long, calculate and plot the the mean life expectancy for each continent over time from 1982 to 2007. Give your plot a title and assign x and y labels. Hint: use the dplyr::group_by() and dplyr::summarise().

ugh, i have no clue


```{r}
byyear <- gap_long %>% 
  filter(obs_type == "lifeExp",
         year > 1980) %>%
  group_by(continent, year) %>% 
  summarise(mean_LE = mean(obs_values)) %>%  #mean_LE new data frame 
  ungroup()


           
ggplot(data = byyear, aes(x = year, y = mean_LE, colour = continent)) +
  geom_line() +
  labs(x = "Year",
       y = "Age (Years)",
       title = "Mean life expectancy") +
  theme_minimal()




```

spread, turn a long dataset into a wide dataset


```{r}
gap_normal <- gap_long %>% 
  spread(key = obs_type, ## converting this column
         value = obs_values) ## creating new columns for each thing in the column to be spread
```


```{r}
## with summarise or summarize

gap_summarised <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap) %>% 
  summarise(max_gdp = max(gdp)) %>% 
  ungroup () %>% 
  arrange(max_gdp)
  
```



7.6 Exercise

    Convert “gap_long” all the way back to gap_wide. Hint: you’ll need to create appropriate labels for all our new variables (time*metric combinations) with the opposite of separate: tidyr::unite().
    
```{r}

#tidyr::unite()

gap_wideagain <- gap_long %>% 
  unite(col = var_names, obs_type, year, sep = "_") %>% # var_names a new column combining obs_type & year
  spread(key = var_names, # then spread
         value = obs_values)

      
```
    
